Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild

### 2020-10-01 IJHCSSubmission/bp_conclusion.tex
removing the past revision blue marks and starting with new ones for this coming revision

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness, and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workers experience stress over time and to
examine the ability of biometrics to predict these
productivity-related aspects. The longitudinal and in-situ placement of
the study support and extend previous work. Based on
twice-collected daily survey responses, we observed that
although participants sometimes saw periods of sustained stress,
they would always return to a baseline stress level for them
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometric data with which we were able to create a model that is able to
predict user stress, sleepiness, and lack of focus with an up to 84.65\% improvement over the baseline.

>>>>

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness, and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workers experience these human aspects over time and to
examine the ability of biometrics to predict these aspects. The longitudinal and in-situ placement of
the study support and extend previous work. Based on
daily collected survey responses, we observed that
although participants sometimes saw periods of sustained stress or sleepiness,
they would always return to their baseline reporting level
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometric data with which we were able to create a model that is able to
predict user stress, sleepiness, and lack of focus with small improvements in accuracy (from 3.1\% to 7.2\%, depending on the aspect in question), and moderate improvements in precision (25.9\% - 52.4\%) in comparison to a stratified random classifier. While the precision and recall scores we report are still low overall, this is a difficult problem to solve and our improvements indicate the potential for future researchers to build upon.

### 2020-10-01 IJHCSSubmission/bp_discussion.tex
removing the past revision blue marks and starting with new ones for this coming revision

%% %\vspace{0.05in}
%% \noindent
%% \textbf{Imbalanced Data:} 
\subsection{Imbalanced Data}
Study participants provided highly
imbalanced data in their survey responses, with most participants only
taking advantage of a subset of the Likert-scale values and the data
points mostly being clustered around the middle of the scale, as can
be seen in Table~\ref{responseDistribution}. While some of the
imbalance is expected due to certain classes, such as `not stressed',
being more common in the workplace, this imbalance also provides
challenges in the training and assessment of a machine learning
classifier, as also found by others, e.g.~\cite{Exler16}. We addressed
this for the training by oversampling in case of few data samples for
the individual models and undersampling in case of a general model
where more data was available. \rev{Rebalancing the dataset using such techniques
is a common and effective practice~\cite{branco2016survey}. Alternative techniques such as SMOTE~\cite{chawla2002smote} exist and can perform better than those we employed, however they are impractical considering the limited amount of data we have to work with. Oversampling the dataset as we did may also lead to an increased risk of model overfitting. However, we believe the benefits of rebalancing the dataset outweigh this possible downside.}

>>>>

%% %\vspace{0.05in}
%% \noindent
%% \textbf{Imbalanced Data:} 
\subsection{Imbalanced Data}
Study participants provided highly
imbalanced data in their survey responses, with most participants only
taking advantage of a subset of the Likert-scale values and the data
points mostly being clustered around the middle of the scale, as can
be seen in Table~\ref{responseDistribution}. While some of the
imbalance is expected due to certain classes, such as `not stressed',
being more common in the workplace, this imbalance also provides
challenges in the training and assessment of a machine learning
classifier, as also found by others, e.g.~\cite{Exler16}. We addressed
this for the training by oversampling in case of few data samples for
the individual models and undersampling in case of a general model
where more data was available. Rebalancing the dataset using such techniques
is a common and effective practice~\cite{branco2016survey}. Alternative techniques such as SMOTE~\cite{chawla2002smote} exist and can perform better than those we employed, however they are impractical considering the limited amount of data we have to work with. Oversampling the dataset as we did may also lead to an increased risk of model overfitting. However, we believe the benefits of rebalancing the dataset outweigh this possible downside.

### 2020-10-01 IJHCSSubmission/bp_relatedWork.tex
removing the past revision blue marks and starting with new ones for this coming revision

The study and prediction approach used in this paper is related
to previous studies of \rev{stress, focus, and awakeness} and studies
using biometrics to predict them. We consider
related work in each of these categories in turn.

@@

\rev{Rather than trying to measure or predict stress, several studies have induced stress and examined its effect on work performance, motivation and others. For instance, Sarsenbayeva et al.~\cite{Sarsenbayeva19} induced stress and studied how it affects mobile interaction, showing that it can reduce completion time and accuracy during target acquisition tasks. Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels. They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits but found no difference in cortisol levels. The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.}

\rev{Stress in the workplace and its effects on service providers has also been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}. These studies considered several factors such as personality traits, career-related goals and attitudes, as well as life outside of work, and examined their correlation with stress levels and burnout.}

\rev{Finally, there is also some work that already used proprietary stress measures for other types of classification. For example, Mirjafari et al.~\cite{Mirjafari19} used a proprietary stress measure by Garmin together with other data collected through the sensing of mobile devices and built machine learning models to differentiate between low and high performing workers.}

@@

\subsection{Focus}
Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
Mark et al.~\cite{mark2014bored} studied \rev{attentional states, including focus, for workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days.} They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon. They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.

Interruptions in the office are a common barriers keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
Such interruptions may include emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}. Interruptions in inopportune times can have negative effects that range from higher error rate and lower overall performance
to an increase in stress and frustration~\cite{bailey2001effects,czerwinski2000instant,mark2008cost}.
External interruptions may cause workers to enter a ``chain of distraction''~\cite{shamsi07}.
This chain is composed by stages of preparation, diversion, resumption and recovery that result in time away from an ongoing task. \rev{Since interruptions can have a large impact on the focus and productivity of office workers, several studies have examined the prediction of interruptibility---the availability for interruptions---using a variety of features, including computer interaction and biometrics~\cite{fogarty2005examining, iqbal2008effects, chen2004using, bailey2008understanding, zuger2015interruptibility, zuger18}. Most of these studies were again conducted for small and controlled tasks over shorter periods of time.}

@@

\subsection{Biometrics}
Our study investigates the prediction of stress, focus, and awakeness using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}\rev{---some of it already mentioned above---}analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
%and found correlations between biometric signals and predicting code quality~\cite{Muller16}, to assess task difficulty in software development~\cite{fritz2014using}, to assess developer's mental load and perceived difficulty while working on small code snippets, and
Biometric signals have also been studied in the context of technology users. For example, Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks. Similarly, biometrics have been used to measure code difficulty by using biometric sensors~\cite{fritz2014using}
and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.

>>>>

The study and prediction approach used in this paper is related
to previous studies of stress, focus, and awakeness and studies
using biometrics to predict them. We consider
related work in each of these categories in turn.

@@

Rather than trying to measure or predict stress, several studies have induced stress and examined its effect on work performance, motivation and others. For instance, Sarsenbayeva et al.~\cite{Sarsenbayeva19} induced stress and studied how it affects mobile interaction, showing that it can reduce completion time and accuracy during target acquisition tasks. Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels. They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits but found no difference in cortisol levels. The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.

Stress in the workplace and its effects on service providers has also been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}. These studies considered several factors such as personality traits, career-related goals and attitudes, as well as life outside of work, and examined their correlation with stress levels and burnout.

Finally, there is also some work that already used proprietary stress measures for other types of classification. For example, Mirjafari et al.~\cite{Mirjafari19} used a proprietary stress measure by Garmin together with other data collected through the sensing of mobile devices and built machine learning models to differentiate between low and high performing workers.

@@

\subsection{Focus}
Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
Mark et al.~\cite{mark2014bored} studied attentional states, including focus, for workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days. They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon. They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.

Interruptions in the office are a common barriers keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
Such interruptions may include emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}. Interruptions in inopportune times can have negative effects that range from higher error rate and lower overall performance
to an increase in stress and frustration~\cite{bailey2001effects,czerwinski2000instant,mark2008cost}.
External interruptions may cause workers to enter a ``chain of distraction''~\cite{shamsi07}.
This chain is composed by stages of preparation, diversion, resumption and recovery that result in time away from an ongoing task. Since interruptions can have a large impact on the focus and productivity of office workers, several studies have examined the prediction of interruptibility---the availability for interruptions---using a variety of features, including computer interaction and biometrics~\cite{fogarty2005examining, iqbal2008effects, chen2004using, bailey2008understanding, zuger2015interruptibility, zuger18}. Most of these studies were again conducted for small and controlled tasks over shorter periods of time.

@@

\subsection{Biometrics}
Our study investigates the prediction of stress, focus, and awakeness using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}---some of it already mentioned above--- analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
%and found correlations between biometric signals and predicting code quality~\cite{Muller16}, to assess task difficulty in software development~\cite{fritz2014using}, to assess developer's mental load and perceived difficulty while working on small code snippets, and
Biometric signals have also been studied in the context of technology users. For example, Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks. Similarly, biometrics have been used to measure code difficulty by using biometric sensors~\cite{fritz2014using}
and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.
